---
title: "Predicting Hourly Precipitation at JFK Airport"
subtitle: "Regression comparison with temporal holdout evaluation"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Objective

This notebook exposes the complete modeling workflow. It sorts observations by time, reserves the latest 20% for final evaluation, tunes ridge regression only on the earlier period, and compares every model on unseen data.

## Load packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readr)
library(tidymodels)
library(tidyr)
```

## Read and validate weather data

```{r}
#| label: read-data
weather <- read_csv(
  here("data", "jfk_weather_snapshot.csv"),
  show_col_types = FALSE,
  col_types = cols(date = col_datetime())
) |>
  arrange(date)

required_columns <- c(
  "date",
  "humidity",
  "temperature_f",
  "station_pressure",
  "wind_speed_mph",
  "precipitation_in"
)

missing_columns <- setdiff(required_columns, names(weather))

if (length(missing_columns) > 0) {
  stop(
    "Weather data is missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

stopifnot(
  anyDuplicated(weather$date) == 0,
  !any(weather$precipitation_in < 0, na.rm = TRUE)
)

weather |>
  summarise(
    observations = n(),
    first_observation = min(date),
    last_observation = max(date),
    missing_predictor_cells = sum(
      is.na(across(where(is.numeric)))
    )
  ) |>
  kable()
```

```{r}
#| label: precipitation-distribution
#| fig-alt: "Histogram showing a large concentration of hourly precipitation observations at zero."
ggplot(weather, aes(x = precipitation_in)) +
  geom_histogram(
    binwidth = 0.01,
    boundary = 0,
    fill = "#2A9D8F",
    color = "white"
  ) +
  coord_cartesian(
    xlim = c(
      0,
      quantile(weather$precipitation_in, 0.995, na.rm = TRUE)
    )
  ) +
  labs(
    title = "Hourly precipitation is strongly zero-inflated",
    x = "Precipitation (inches)",
    y = "Observations"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Exploratory analysis

The predictors operate on different scales. Faceted histograms reveal skewness, concentration, and potential measurement artifacts before modeling.

```{r}
#| label: predictor-distributions
#| fig-alt: "Faceted histograms showing the distributions of humidity, temperature, station pressure, and wind speed."
weather |>
  select(
    humidity,
    temperature_f,
    station_pressure,
    wind_speed_mph
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "predictor",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(
    bins = 30,
    fill = "#2A9D8F",
    color = "white"
  ) +
  facet_wrap(
    vars(predictor),
    scales = "free",
    ncol = 2
  ) +
  labs(
    title = "Weather predictor distributions",
    x = NULL,
    y = "Observations"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

Daily aggregation makes wet periods visible without drawing thousands of hourly points.

```{r}
#| label: precipitation-time-series
#| fig-alt: "Daily precipitation totals at JFK through the observed period."
daily_precipitation <- weather |>
  mutate(day = as.Date(date)) |>
  group_by(day) |>
  summarise(
    precipitation_in = sum(precipitation_in, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  daily_precipitation,
  aes(x = day, y = precipitation_in)
) +
  geom_col(fill = "#457B9D", width = 1) +
  labs(
    title = "Daily precipitation varies in short, irregular episodes",
    x = "Date",
    y = "Daily precipitation (inches)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Create training, tuning, and test sets

```{r}
#| label: data-split
set.seed(2026)

weather_split <- initial_time_split(weather, prop = 0.8)
training_data <- training(weather_split)
testing_data <- testing(weather_split)
resamples <- vfold_cv(training_data, v = 5)

tibble(
  partition = c("Training and tuning", "Final test"),
  observations = c(nrow(training_data), nrow(testing_data)),
  first_date = c(min(training_data$date), min(testing_data$date)),
  last_date = c(max(training_data$date), max(testing_data$date))
) |>
  kable()
```

## Mean baseline

```{r}
#| label: baseline-model
baseline_predictions <- testing_data |>
  transmute(
    precipitation_in,
    .pred = mean(training_data$precipitation_in)
  )
```

## Shared preprocessing recipe

Median imputation and normalization are estimated inside each workflow, so the final test period cannot influence preprocessing.

```{r}
#| label: preprocessing-recipe
weather_recipe <- recipe(
  precipitation_in ~
    humidity +
    temperature_f +
    station_pressure +
    wind_speed_mph,
  data = training_data
) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Linear regression

```{r}
#| label: linear-model
linear_workflow <- workflow() |>
  add_recipe(weather_recipe) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

linear_fit <- fit(linear_workflow, data = training_data)

linear_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(linear_fit, new_data = testing_data)
)
```

## Ridge regression with cross-validation

```{r}
#| label: ridge-model
ridge_workflow <- workflow() |>
  add_recipe(weather_recipe) |>
  add_model(
    linear_reg(
      penalty = tune(),
      mixture = 0
    ) |>
      set_engine("glmnet")
  )

ridge_grid <- grid_regular(
  penalty(range = c(-5, 0)),
  levels = 8
)

ridge_results <- tune_grid(
  ridge_workflow,
  resamples = resamples,
  grid = ridge_grid,
  metrics = metric_set(rmse, mae, rsq_trad)
)

best_penalty <- select_best(ridge_results, metric = "rmse")
best_penalty |> kable(digits = 5)

ridge_fit <- ridge_workflow |>
  finalize_workflow(best_penalty) |>
  fit(data = training_data)

ridge_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(ridge_fit, new_data = testing_data)
)
```

The tuning profile shows whether the selected penalty lies in a stable region or at the edge of the search grid.

```{r}
#| label: ridge-tuning-curve
#| fig-alt: "Cross-validation RMSE across ridge penalty values."
ridge_tuning <- ridge_results |>
  collect_metrics() |>
  filter(.metric == "rmse")

ggplot(
  ridge_tuning,
  aes(x = penalty, y = mean)
) +
  geom_ribbon(
    aes(
      ymin = mean - std_err,
      ymax = mean + std_err
    ),
    fill = "#A8DADC",
    alpha = 0.6
  ) +
  geom_line(color = "#1D3557", linewidth = 0.9) +
  geom_point(color = "#1D3557", size = 2) +
  scale_x_log10() +
  labs(
    title = "Ridge penalty selected through cross-validation",
    x = "Penalty (log scale)",
    y = "Cross-validation RMSE"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Polynomial regression

```{r}
#| label: polynomial-model
polynomial_recipe <- weather_recipe |>
  step_poly(
    temperature_f,
    humidity,
    station_pressure,
    wind_speed_mph,
    degree = 2
  )

polynomial_workflow <- workflow() |>
  add_recipe(polynomial_recipe) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

polynomial_fit <- fit(
  polynomial_workflow,
  data = training_data
)

polynomial_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(polynomial_fit, new_data = testing_data)
)
```

## Final holdout performance

```{r}
#| label: holdout-metrics
baseline_metrics <- bind_rows(
  rmse(baseline_predictions, precipitation_in, .pred),
  mae(baseline_predictions, precipitation_in, .pred),
  rsq_trad(baseline_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Mean baseline")

linear_metrics <- bind_rows(
  rmse(linear_predictions, precipitation_in, .pred),
  mae(linear_predictions, precipitation_in, .pred),
  rsq_trad(linear_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Linear regression")

ridge_metrics <- bind_rows(
  rmse(ridge_predictions, precipitation_in, .pred),
  mae(ridge_predictions, precipitation_in, .pred),
  rsq_trad(ridge_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Ridge regression")

polynomial_metrics <- bind_rows(
  rmse(polynomial_predictions, precipitation_in, .pred),
  mae(polynomial_predictions, precipitation_in, .pred),
  rsq_trad(polynomial_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Polynomial regression")

holdout_metrics <- bind_rows(
  baseline_metrics,
  linear_metrics,
  ridge_metrics,
  polynomial_metrics
) |>
  select(model, .metric, .estimate)

metrics_table <- holdout_metrics |>
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) |>
  arrange(rmse)

metrics_table |>
  kable(digits = 4)
```

RMSE emphasizes larger misses, MAE describes typical absolute error, and R² measures improvement relative to a constant mean on the same holdout.

## Observed versus predicted

```{r}
#| label: observed-predicted
#| fig-height: 7
#| fig-alt: "Faceted observed-versus-predicted plots for wet hours across the four precipitation models."
prediction_comparison <- bind_rows(
  baseline_predictions |>
    mutate(model = "Mean baseline"),
  linear_predictions |>
    mutate(model = "Linear regression"),
  ridge_predictions |>
    mutate(model = "Ridge regression"),
  polynomial_predictions |>
    mutate(model = "Polynomial regression")
) |>
  mutate(
    model = factor(
      model,
      levels = c(
        "Mean baseline",
        "Linear regression",
        "Ridge regression",
        "Polynomial regression"
      )
    )
  )

wet_prediction_comparison <- prediction_comparison |>
  filter(precipitation_in > 0)

ggplot(
  wet_prediction_comparison,
  aes(x = precipitation_in, y = .pred)
) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "#E76F51",
    linewidth = 0.8
  ) +
  geom_point(alpha = 0.35, color = "#2A9D8F") +
  facet_wrap(
    vars(model),
    scales = "free_y",
    ncol = 2
  ) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  labs(
    title = "Observed and predicted precipitation during wet holdout hours",
    x = "Observed precipitation (inches)",
    y = "Predicted precipitation (inches)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Residual comparison

```{r}
#| label: residual-comparison
#| fig-alt: "Faceted residual plots for the four precipitation models."
residual_comparison <- prediction_comparison |>
  mutate(residual = precipitation_in - .pred)

ggplot(
  residual_comparison,
  aes(x = .pred, y = residual)
) +
  geom_hline(yintercept = 0, color = "#555555") +
  geom_point(alpha = 0.3, color = "#457B9D") +
  geom_smooth(se = FALSE, color = "#E76F51") +
  facet_wrap(vars(model), scales = "free_x", ncol = 2) +
  labs(
    title = "Residual structure reveals systematic underprediction",
    x = "Predicted precipitation (inches)",
    y = "Observed minus predicted"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Conclusion

- Polynomial regression achieves the lowest holdout RMSE (`r sprintf("%.4f", metrics_table$rmse[metrics_table$model == "Polynomial regression"])`), a `r sprintf("%.1f", 100 * (metrics_table$rmse[metrics_table$model == "Mean baseline"] - metrics_table$rmse[metrics_table$model == "Polynomial regression"]) / metrics_table$rmse[metrics_table$model == "Mean baseline"])`% reduction relative to the mean baseline.
- Its holdout R² is only `r sprintf("%.3f", metrics_table$rsq_trad[metrics_table$model == "Polynomial regression"])`, so the available contemporaneous predictors explain little of the temporal variation in precipitation.
- The ridge penalty selected by five-fold cross-validation is `r format(best_penalty$penalty[[1]], digits = 3, scientific = TRUE)`. Ridge and linear regression produce nearly identical RMSE, indicating limited benefit from regularization with only four predictors.
- The mean baseline has the lowest MAE (`r sprintf("%.4f", metrics_table$mae[metrics_table$model == "Mean baseline"])`) despite the worst RMSE. This metric reversal is consistent with zero inflation: predicting near zero performs well for typical hours, while all models underpredict uncommon heavy-rain events.
- Operational improvement would require lagged precipitation, radar, cloud, and atmospheric predictors or a two-stage occurrence-and-amount model.

See the [Spanish version](es/index.qmd).
