---
title: "Forecasting Hourly Precipitation at JFK Airport"
subtitle: "A calibrated two-stage model with temporal validation"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Objective

Hourly precipitation is a zero-inflated target: most observations are dry, while a small number of storms account for the largest errors. This notebook therefore answers two questions separately: **will it rain?** and, conditional on rain, **how much?**

Every transformation, model, and evaluation step is visible below. The forecast uses contemporaneous weather readings plus the previous precipitation observation; it never uses future information.

## Packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(tidyverse)
library(tidymodels)
library(here)
library(knitr)
stopifnot(requireNamespace("xgboost", quietly = TRUE))
```

## Read the reproducible snapshot

```{r}
#| label: read-data
weather <- read_csv(
  here("data", "jfk_weather_snapshot.csv"),
  show_col_types = FALSE,
  col_types = cols(date = col_datetime())
) |>
  arrange(date)
```

```{r}
#| label: validate-data
required_columns <- c(
  "date", "humidity", "temperature_f", "station_pressure",
  "wind_speed_mph", "precipitation_in"
)

stopifnot(
  all(required_columns %in% names(weather)),
  anyDuplicated(weather$date) == 0,
  !any(weather$precipitation_in < 0, na.rm = TRUE)
)
```

## Engineer available-at-prediction-time features

```{r}
#| label: feature-engineering
weather <- weather |>
  mutate(
    rain = factor(if_else(precipitation_in > 0, "rain", "dry"),
      levels = c("rain", "dry")),
    precipitation_lag1 = lag(precipitation_in, default = 0),
    rain_lag1 = lag(as.integer(precipitation_in > 0), default = 0),
    gap_hours = as.numeric(difftime(date, lag(date), units = "hours")),
    gap_hours = replace_na(gap_hours, median(gap_hours, na.rm = TRUE)),
    month_sin = sin(2 * pi * month(date) / 12),
    month_cos = cos(2 * pi * month(date) / 12),
    hour_sin = sin(2 * pi * hour(date) / 24),
    hour_cos = cos(2 * pi * hour(date) / 24)
  )
```

```{r}
#| label: data-summary
weather |>
  summarise(
    observations = n(),
    first_observation = min(date),
    last_observation = max(date),
    wet_hours = sum(rain == "rain"),
    wet_share = mean(rain == "rain")
  ) |>
  kable(digits = 3)
```

## Explore the target

```{r}
#| label: precipitation-distribution
#| fig-alt: "Histogram showing that most hourly precipitation observations are zero."
ggplot(weather, aes(x = precipitation_in)) +
  geom_histogram(binwidth = 0.01, boundary = 0,
    fill = "#2A9D8F", color = "white") +
  coord_cartesian(xlim = c(0, quantile(weather$precipitation_in, 0.995))) +
  labs(
    title = "Hourly precipitation is strongly zero-inflated",
    x = "Precipitation (inches)", y = "Observations"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r}
#| label: daily-series
#| fig-alt: "Daily precipitation totals at JFK over time."
weather |>
  mutate(day = as.Date(date)) |>
  group_by(day) |>
  summarise(precipitation_in = sum(precipitation_in), .groups = "drop") |>
  ggplot(aes(x = day, y = precipitation_in)) +
  geom_col(fill = "#457B9D", width = 1) +
  labs(
    title = "Rain arrives in short, irregular episodes",
    x = "Date", y = "Daily precipitation (inches)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Preserve the temporal order

The earliest 70% is used for model selection, the next 10% for calibration, and the latest 20% only once for final evaluation.

```{r}
#| label: temporal-splits
set.seed(2026)
final_split <- initial_time_split(weather, prop = 0.8)
development_data <- training(final_split)
testing_data <- testing(final_split)

validation_split <- initial_time_split(development_data, prop = 0.875)
modeling_data <- training(validation_split)
validation_data <- testing(validation_split)
```

```{r}
#| label: partition-table
bind_rows(
  modeling_data |> summarise(partition = "Modeling", n = n(), first = min(date), last = max(date)),
  validation_data |> summarise(partition = "Calibration", n = n(), first = min(date), last = max(date)),
  testing_data |> summarise(partition = "Final test", n = n(), first = min(date), last = max(date))
) |>
  kable()
```

```{r}
#| label: temporal-resamples
class_resamples <- rolling_origin(
  modeling_data,
  initial = 1500,
  assess = 325,
  cumulative = TRUE,
  skip = 324
)

class_resamples
```

## Baselines

```{r}
#| label: baselines
baseline_mean <- mean(development_data$precipitation_in)

baseline_predictions <- testing_data |>
  transmute(
    date, precipitation_in,
    `Dry baseline` = 0,
    `Historical mean` = baseline_mean
  )
```

### Previous single-stage benchmark

```{r}
#| label: polynomial-recipe
polynomial_recipe <- recipe(
  precipitation_in ~ humidity + temperature_f +
    station_pressure + wind_speed_mph,
  data = development_data
) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_poly(all_numeric_predictors(), degree = 2)
```

```{r}
#| label: polynomial-model
polynomial_fit <- workflow() |>
  add_recipe(polynomial_recipe) |>
  add_model(linear_reg() |> set_engine("lm")) |>
  fit(data = development_data)

polynomial_predictions <- predict(polynomial_fit, testing_data) |>
  pull(.pred)
```

## Stage 1: rain occurrence

### Classification recipe

```{r}
#| label: occurrence-recipe
occurrence_recipe <- recipe(
  rain ~ humidity + temperature_f + station_pressure + wind_speed_mph +
    precipitation_lag1 + rain_lag1 + gap_hours + month_sin + month_cos +
    hour_sin + hour_cos,
  data = modeling_data
) |>
  step_impute_median(all_numeric_predictors())
```

### Classification model

```{r}
#| label: occurrence-model
occurrence_spec <- boost_tree(
  trees = 600,
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune()
) |>
  set_mode("classification") |>
  set_engine("xgboost")
```

### Classification grid

```{r}
#| label: occurrence-grid
predictor_names <- c(
  "humidity", "temperature_f", "station_pressure", "wind_speed_mph",
  "precipitation_lag1", "rain_lag1", "gap_hours", "month_sin",
  "month_cos", "hour_sin", "hour_cos"
)

occurrence_grid <- grid_space_filling(
  tree_depth(c(2L, 7L)), min_n(c(5L, 35L)), learn_rate(c(-3, -0.7)),
  loss_reduction(), sample_prop(c(0.65, 1)),
  finalize(mtry(), modeling_data |> select(all_of(predictor_names))),
  size = 12
)
```

### Temporal tuning

```{r}
#| label: occurrence-tuning
occurrence_results <- workflow() |>
  add_recipe(occurrence_recipe) |>
  add_model(occurrence_spec) |>
  tune_grid(
    resamples = class_resamples,
    grid = occurrence_grid,
    metrics = metric_set(pr_auc, roc_auc, mn_log_loss)
  )

best_occurrence <- select_best(occurrence_results, metric = "pr_auc")
best_occurrence |> kable(digits = 4)
```

### Fit the selected classifier

```{r}
#| label: occurrence-fit
occurrence_workflow <- workflow() |>
  add_recipe(occurrence_recipe) |>
  add_model(occurrence_spec) |>
  finalize_workflow(best_occurrence)

occurrence_fit <- occurrence_workflow |>
  fit(data = modeling_data)
```

## Stage 2: precipitation amount

The amount model sees only wet observations. This prevents thousands of dry hours from overwhelming the storm-size signal.

```{r}
#| label: wet-training-data
wet_modeling_data <- modeling_data |>
  filter(rain == "rain")

amount_resamples <- rolling_origin(
  wet_modeling_data,
  initial = 140,
  assess = 40,
  cumulative = TRUE,
  skip = 39
)
```

### Amount recipe

```{r}
#| label: amount-recipe
amount_recipe <- recipe(
  precipitation_in ~ humidity + temperature_f + station_pressure +
    wind_speed_mph + precipitation_lag1 + rain_lag1 + gap_hours +
    month_sin + month_cos + hour_sin + hour_cos,
  data = wet_modeling_data
) |>
  step_impute_median(all_numeric_predictors())
```

### Amount model

```{r}
#| label: amount-model
amount_spec <- boost_tree(
  trees = 600,
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune()
) |>
  set_mode("regression") |>
  set_engine("xgboost")
```

### Amount grid and tuning

```{r}
#| label: amount-grid
amount_grid <- grid_space_filling(
  tree_depth(c(2L, 6L)), min_n(c(3L, 25L)), learn_rate(c(-3, -0.7)),
  loss_reduction(), sample_prop(c(0.65, 1)),
  finalize(mtry(), wet_modeling_data |> select(all_of(predictor_names))),
  size = 12
)
```

```{r}
#| label: amount-tuning
amount_results <- workflow() |>
  add_recipe(amount_recipe) |>
  add_model(amount_spec) |>
  tune_grid(
    resamples = amount_resamples,
    grid = amount_grid,
    metrics = metric_set(rmse, mae)
  )

best_amount <- select_best(amount_results, metric = "rmse")
best_amount |> kable(digits = 4)
```

### Fit the selected amount model

```{r}
#| label: amount-fit
amount_workflow <- workflow() |>
  add_recipe(amount_recipe) |>
  add_model(amount_spec) |>
  finalize_workflow(best_amount)

amount_fit <- amount_workflow |>
  fit(data = wet_modeling_data)
```

## Calibrate without touching the final test

```{r}
#| label: validation-predictions
validation_predictions <- validation_data |>
  select(date, rain, precipitation_in) |>
  bind_cols(predict(occurrence_fit, validation_data, type = "prob")) |>
  bind_cols(predict(amount_fit, validation_data)) |>
  rename(amount_raw = .pred) |>
  mutate(
    amount_raw = pmax(amount_raw, 0),
    logit_probability = qlogis(pmin(pmax(.pred_rain, 0.001), 0.999))
  )
```

```{r}
#| label: probability-calibration
probability_calibration <- glm(
  I(rain == "rain") ~ logit_probability,
  data = validation_predictions,
  family = binomial()
)

validation_predictions <- validation_predictions |>
  mutate(rain_probability = predict(
    probability_calibration, newdata = validation_predictions,
    type = "response"
  ))
```

```{r}
#| label: amount-calibration
amount_scale <- validation_predictions |>
  filter(rain == "rain") |>
  summarise(scale = sum(precipitation_in * amount_raw) / sum(amount_raw^2)) |>
  pull(scale)

amount_scale
```

```{r}
#| label: threshold-selection
threshold_results <- crossing(
  threshold = seq(0.05, 0.60, by = 0.01),
  validation_predictions
) |>
  mutate(.pred = if_else(
    rain_probability >= threshold, amount_raw * amount_scale, 0
  )) |>
  group_by(threshold) |>
  summarise(rmse = sqrt(mean((precipitation_in - .pred)^2)), .groups = "drop")

best_threshold <- threshold_results |>
  slice_min(rmse, n = 1, with_ties = FALSE) |>
  pull(threshold)
```

## Refit on all development data

```{r}
#| label: final-fits
final_occurrence_fit <- occurrence_workflow |>
  fit(data = development_data)

final_amount_fit <- amount_workflow |>
  fit(data = development_data |> filter(rain == "rain"))
```

```{r}
#| label: holdout-predictions
holdout_predictions <- testing_data |>
  select(date, rain, precipitation_in) |>
  bind_cols(predict(final_occurrence_fit, testing_data, type = "prob")) |>
  bind_cols(predict(final_amount_fit, testing_data)) |>
  rename(amount_raw = .pred) |>
  mutate(
    amount_raw = pmax(amount_raw, 0),
    logit_probability = qlogis(pmin(pmax(.pred_rain, 0.001), 0.999)),
    rain_probability = predict(probability_calibration,
      newdata = pick(everything()), type = "response"),
    amount = amount_raw * amount_scale
  )
```

```{r}
#| label: combined-predictions
holdout_predictions <- holdout_predictions |>
  mutate(
    `Hurdle model` = if_else(
      rain_probability >= best_threshold, amount, 0
    ),
    `Expected precipitation` = rain_probability * amount,
    `Single-stage polynomial` = polynomial_predictions,
    `Dry baseline` = 0,
    `Historical mean` = baseline_mean
  )
```

## Final holdout performance

```{r}
#| label: regression-metrics
regression_metrics <- holdout_predictions |>
  select(precipitation_in, `Dry baseline`, `Historical mean`,
    `Single-stage polynomial`, `Hurdle model`, `Expected precipitation`) |>
  pivot_longer(-precipitation_in, names_to = "model", values_to = ".pred") |>
  group_by(model) |>
  summarise(
    rmse = sqrt(mean((precipitation_in - .pred)^2)),
    mae = mean(abs(precipitation_in - .pred)),
    wet_rmse = sqrt(mean((precipitation_in[precipitation_in > 0] -
      .pred[precipitation_in > 0])^2)),
    .groups = "drop"
  ) |>
  arrange(rmse)

regression_metrics |> kable(digits = 4)
```

```{r}
#| label: occurrence-metrics
occurrence_metrics <- holdout_predictions |>
  summarise(
    pr_auc = pr_auc_vec(rain, rain_probability, event_level = "first"),
    roc_auc = roc_auc_vec(rain, rain_probability, event_level = "first"),
    brier_score = brier_class_vec(rain, rain_probability,
      event_level = "first")
  )

occurrence_metrics |> kable(digits = 3)
```

## Inspect predictions over time

```{r}
#| label: holdout-series-data
holdout_daily <- holdout_predictions |>
  mutate(day = as.Date(date)) |>
  group_by(day) |>
  summarise(
    Observed = sum(precipitation_in),
    Predicted = sum(`Expected precipitation`),
    .groups = "drop"
  ) |>
  pivot_longer(c(Observed, Predicted), names_to = "series", values_to = "inches")
```

```{r}
#| label: holdout-series
#| fig-alt: "Observed and predicted daily precipitation totals in the final test period."
ggplot(holdout_daily, aes(x = day, y = inches, color = series)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(Observed = "#264653", Predicted = "#E76F51")) +
  labs(
    title = "Observed and predicted precipitation - final holdout",
    x = "Date", y = "Daily precipitation (inches)", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")
```

```{r}
#| label: calibration-plot
#| fig-alt: "Rain probability calibration by prediction decile."
holdout_predictions |>
  mutate(probability_bin = ntile(rain_probability, 10)) |>
  group_by(probability_bin) |>
  summarise(
    predicted = mean(rain_probability),
    observed = mean(rain == "rain"),
    .groups = "drop"
  ) |>
  ggplot(aes(x = predicted, y = observed)) +
  geom_abline(slope = 1, intercept = 0, color = "#777777") +
  geom_point(size = 2.5, color = "#2A9D8F") +
  labs(title = "Calibrated rain probabilities", x = "Predicted", y = "Observed") +
  theme_minimal(base_size = 12)
```

## Conclusion

```{r}
#| label: conclusion-values
best_model <- regression_metrics |> slice_min(rmse, n = 1)
mean_baseline_rmse <- regression_metrics |>
  filter(model == "Historical mean") |>
  pull(rmse)
rmse_improvement <- 100 * (mean_baseline_rmse - best_model$rmse) /
  mean_baseline_rmse
```

- The best final prediction is **`r best_model$model`**, with RMSE **`r sprintf("%.4f", best_model$rmse)`** and a **`r sprintf("%.1f", rmse_improvement)`%** reduction relative to the historical-mean baseline.
- The occurrence model reaches PR AUC **`r sprintf("%.3f", occurrence_metrics$pr_auc)`** and ROC AUC **`r sprintf("%.3f", occurrence_metrics$roc_auc)`** on the untouched final period.
- The calibration threshold (**`r sprintf("%.2f", best_threshold)`**) and amount correction (**`r sprintf("%.2f", amount_scale)`**) were learned from the middle 10%, not from the final test.
- The model is reproducible and leakage-aware, but it is not a production nowcast: radar, cloud cover, and denser hourly history would be needed for that claim.

See the [Spanish version](es/index.qmd).
