---
title: "Predicting Hourly Precipitation at JFK Airport"
subtitle: "Regression comparison with temporal holdout evaluation"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

## Objective

This notebook exposes the complete modeling workflow. It sorts observations by time, reserves the latest 20% for final evaluation, tunes ridge regression only on the earlier period, and compares every model on unseen data.

## Load packages

```{r}
#| label: packages
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readr)
library(tidymodels)
library(tidyr)
```

## Read and validate weather data

```{r}
#| label: read-data
weather <- read_csv(
  here("data", "jfk_weather_snapshot.csv"),
  show_col_types = FALSE,
  col_types = cols(date = col_datetime())
) |>
  arrange(date)

required_columns <- c(
  "date",
  "humidity",
  "temperature_f",
  "station_pressure",
  "wind_speed_mph",
  "precipitation_in"
)

missing_columns <- setdiff(required_columns, names(weather))

if (length(missing_columns) > 0) {
  stop(
    "Weather data is missing columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

stopifnot(
  anyDuplicated(weather$date) == 0,
  !any(weather$precipitation_in < 0, na.rm = TRUE)
)

weather |>
  summarise(
    observations = n(),
    first_observation = min(date),
    last_observation = max(date),
    missing_predictor_cells = sum(
      is.na(across(where(is.numeric)))
    )
  ) |>
  kable()
```

```{r}
#| label: precipitation-distribution
#| fig-alt: "Histogram showing a large concentration of hourly precipitation observations at zero."
ggplot(weather, aes(x = precipitation_in)) +
  geom_histogram(
    binwidth = 0.01,
    boundary = 0,
    fill = "#2A9D8F",
    color = "white"
  ) +
  coord_cartesian(
    xlim = c(
      0,
      quantile(weather$precipitation_in, 0.995, na.rm = TRUE)
    )
  ) +
  labs(
    title = "Hourly precipitation is strongly zero-inflated",
    x = "Precipitation (inches)",
    y = "Observations"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Create training, tuning, and test sets

```{r}
#| label: data-split
set.seed(2026)

weather_split <- initial_time_split(weather, prop = 0.8)
training_data <- training(weather_split)
testing_data <- testing(weather_split)
resamples <- vfold_cv(training_data, v = 5)

tibble(
  partition = c("Training and tuning", "Final test"),
  observations = c(nrow(training_data), nrow(testing_data)),
  first_date = c(min(training_data$date), min(testing_data$date)),
  last_date = c(max(training_data$date), max(testing_data$date))
) |>
  kable()
```

## Mean baseline

```{r}
#| label: baseline-model
baseline_predictions <- testing_data |>
  transmute(
    precipitation_in,
    .pred = mean(training_data$precipitation_in)
  )
```

## Shared preprocessing recipe

Median imputation and normalization are estimated inside each workflow, so the final test period cannot influence preprocessing.

```{r}
#| label: preprocessing-recipe
weather_recipe <- recipe(
  precipitation_in ~
    humidity +
    temperature_f +
    station_pressure +
    wind_speed_mph,
  data = training_data
) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Linear regression

```{r}
#| label: linear-model
linear_workflow <- workflow() |>
  add_recipe(weather_recipe) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

linear_fit <- fit(linear_workflow, data = training_data)

linear_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(linear_fit, new_data = testing_data)
)
```

## Ridge regression with cross-validation

```{r}
#| label: ridge-model
ridge_workflow <- workflow() |>
  add_recipe(weather_recipe) |>
  add_model(
    linear_reg(
      penalty = tune(),
      mixture = 0
    ) |>
      set_engine("glmnet")
  )

ridge_grid <- grid_regular(
  penalty(range = c(-5, 0)),
  levels = 8
)

ridge_results <- tune_grid(
  ridge_workflow,
  resamples = resamples,
  grid = ridge_grid,
  metrics = metric_set(rmse, mae, rsq_trad)
)

best_penalty <- select_best(ridge_results, metric = "rmse")
best_penalty |> kable(digits = 5)

ridge_fit <- ridge_workflow |>
  finalize_workflow(best_penalty) |>
  fit(data = training_data)

ridge_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(ridge_fit, new_data = testing_data)
)
```

## Polynomial regression

```{r}
#| label: polynomial-model
polynomial_recipe <- weather_recipe |>
  step_poly(
    temperature_f,
    humidity,
    station_pressure,
    wind_speed_mph,
    degree = 2
  )

polynomial_workflow <- workflow() |>
  add_recipe(polynomial_recipe) |>
  add_model(
    linear_reg() |>
      set_engine("lm")
  )

polynomial_fit <- fit(
  polynomial_workflow,
  data = training_data
)

polynomial_predictions <- bind_cols(
  testing_data |> select(precipitation_in),
  predict(polynomial_fit, new_data = testing_data)
)
```

## Final holdout performance

```{r}
#| label: holdout-metrics
baseline_metrics <- bind_rows(
  rmse(baseline_predictions, precipitation_in, .pred),
  mae(baseline_predictions, precipitation_in, .pred),
  rsq_trad(baseline_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Mean baseline")

linear_metrics <- bind_rows(
  rmse(linear_predictions, precipitation_in, .pred),
  mae(linear_predictions, precipitation_in, .pred),
  rsq_trad(linear_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Linear regression")

ridge_metrics <- bind_rows(
  rmse(ridge_predictions, precipitation_in, .pred),
  mae(ridge_predictions, precipitation_in, .pred),
  rsq_trad(ridge_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Ridge regression")

polynomial_metrics <- bind_rows(
  rmse(polynomial_predictions, precipitation_in, .pred),
  mae(polynomial_predictions, precipitation_in, .pred),
  rsq_trad(polynomial_predictions, precipitation_in, .pred)
) |>
  mutate(model = "Polynomial regression")

holdout_metrics <- bind_rows(
  baseline_metrics,
  linear_metrics,
  ridge_metrics,
  polynomial_metrics
) |>
  select(model, .metric, .estimate)

holdout_metrics |>
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) |>
  arrange(rmse) |>
  kable(digits = 4)
```

RMSE emphasizes larger misses, MAE describes typical absolute error, and R² measures improvement relative to a constant mean on the same holdout.

## Conclusion

The visible code demonstrates temporal splitting, leakage-safe preprocessing, cross-validation, hyperparameter selection, prediction, and final evaluation. See the [Spanish version](es/index.qmd).
