---
title: "Predicting Hourly Precipitation at JFK Airport"
subtitle: "Regression comparison with temporal holdout evaluation"
author: "Miguelangel Palma Rojas"
date: last-modified
lang: en
---

```{r}
#| label: setup
#| include: false
source(here::here("R", "analysis.R"))
weather <- read_weather_snapshot()
model_results <- fit_weather_models(weather)
```

## Objective

This notebook predicts hourly precipitation from humidity, temperature, station pressure, and wind speed. It corrects the original evaluation by sorting observations by time, reserving the latest 20% as a final test set, and using cross-validation only inside the earlier training period.

## Data preparation

The snapshot contains `r nrow(weather)` validated observations. NOAA trace values (`T`) are documented and converted to zero inches; malformed numeric suffixes are removed; records without a timestamp or outcome are excluded.

```{r}
weather |>
  summarise(
    first_observation = min(date),
    last_observation = max(date),
    missing_predictor_cells = sum(is.na(across(where(is.numeric))))
  ) |>
  knitr::kable()
```

```{r}
#| fig-alt: "Histogram showing a large concentration of hourly precipitation observations at zero."
plot_precipitation_distribution(weather, "en")
```

## Models

- **Mean baseline:** establishes the minimum useful benchmark.
- **Linear regression:** interpretable additive relationships.
- **Ridge regression:** regularizes correlated weather predictors; its penalty is selected by cross-validation.
- **Polynomial regression:** allows smooth curvature while preserving a transparent regression structure.

## Final holdout performance

```{r}
model_results$metrics |>
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(rmse) |>
  knitr::kable(digits = 4)
```

Every metric above uses only the final time-ordered test set. RMSE emphasizes larger misses, MAE describes typical absolute error, and R² measures improvement relative to a constant mean under the same holdout.

## Interpretation

Precipitation is zero-inflated and difficult to predict with a small set of contemporaneous measurements. A regression with lower error than the baseline is useful evidence, but operational forecasting would require lagged radar, cloud, and atmospheric variables.

## Conclusion

The primary improvement is trustworthy evaluation. Model selection occurs before the final test period, and the notebook reports several complementary metrics instead of training error.

